Retirement Consumption: Evidence from UK Pension Reform

### 2023-07-26 Texfiles/Draft1.tex
Latest edits

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \cite{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts - this phenomenon has been termed the "annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and a function of wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year. Subsequently
the number of annuities sold in the UK dropped precipitously.

In this paper, I first use the policy reform as a discontinuity to measure the
impact of forced annuitisation on the consumption levels of individuals in the
first few years of retirement. Given that the reform was implemented suddenly
and without advanced notice before the Spring 2014 budget, I claim that
individuals who retired in 2015, 2016 or 2017 are otherwise similar to those who
retired in 2011, 2012 and 2013 except for the fact the later retirees do not
need to annuitise their DC pension pots. Therefore, I can use a regression
discontinuity to identify the effect that forced annuitisation has on the
consumption of retirees.

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. The annuity puzzle was first identified by
\cite{yaari_65} and is a classed as a problem since standard economic theory
cannot explain the lack of annuitisation that is observed in the real world. Two
possible explanations are bequests and life expectancy: individuals want to
leave an estate for their heirs to inherit and they cannot do this if they
annuitise their wealth; or individuals may believe they will not live as long as
annuity providers think they will and therefore annuities that are on the market
do not appear to be good value. Depending on the reason for the lack of
annuitisation in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life
expectancy their consumption should increase after the reform. If, on the other
hand, individuals do not annuitise \ because of a bequest motive, consumption
should not change much as a result of the reform.

I will solve lifecycle models (\textbf{maybe add something about lifecycle
    models here}) for both of these cases and simulate consumption decisions with
and without forced annuitisation. I will then use a variety of empirical models
to measure the consumption change in early retirement that resulted from the
policy reform. The sign and magnitude of this change will be indicative of the
mechanism causing the annuitisation problem.

The importance of retirement policy to individuals in the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS) and for every 1000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\cite{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \cite{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals will require expensive end of life care. On the
other hand, if retirees over-save and do not spend, the economy may be
dynamically inefficient whereby the capital stock is too high because
individuals save too much for retirement and for bequests. In this case
consumption per capita could be increased if savings were decreased.

\subsection{Literature review}
My paper builds on three main strands of literature: that of the annuity puzzle,
exploring reasons that some retirees choose not to annuitise; the retirement
saving puzzle, which seeks to explain why retirees drawdown assets slowly, ; and
lifecycle models, whereby the above problems are explained using a tractable
model of human decision making.

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. And more recently
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions.
Since then there has been much literature discussing possible reasons that
people do not annuitise. \cite{finkelstein_porteba_2002,
    finkelstein_porteba_2004} for the UK market and \cite{mitchell_et_al_1999} for
the US market and  find evidence of adverse selection, thereby making the
'money's worth' of annuities lower for the general population as opposed to the
population of annuitants. However, they also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from actuarily fair.

\cite{friedman_warshawsky_qje_1990} show that annuitisation decisions can be
fully explained by a mixture of bequest motives and actuarily unfair annuities.
They solve an augmented life-cycle model with a range of parameters on how
severe the rate of return is on the annuity versus market rates. For plausible
values they find that individuals would optimally not annuitise much wealth.
Similarly to \cite{finkelstein_porteba_2004},
\cite{friedman_warshawsky_chicago_1988} show that there is a significant
difference between the life expectancy of annuitants and the general population
in the American annuity market but this cannot fully explain the annuitisation
puzzle. Only when bequest motives are added to the model can annuitisation rates
be rationalized.

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks therefore match other important aspects of the
lifecycle model such as how much individuals actually bequest and how rich
individuals are when they bequest. Also, \cite{odea_sturrock_rest_2023} find that subjective life expectancies can
explain a large part of the annuitisation puzzle especially if agents are less
risk averse.

\cite{lockwood_aer_2018}

\cite{vidalmelia_lejarragagarcia_munich_2004} have some interesting results.
Need to talk about that.

\textbf{There are some more papers I will include here.}

@@

The main dataset I use is the English Longitudinal Study of Ageing (ELSA)
\cite{main_elsa_citation}. ELSA picks individuals over the age of 50 to survey
every two years until death. If individuals leave the survey, ELSA replaces them
so that it is representative of the over 50 population in the UK. Individuals
are asked a range of questions relating to their income and wealth as well as
expectations about the future. One benefit of using ELSA is that it includes
data on pension types for individuals who are working. Therefore, I can
differentiate between individuals who have a DC pension and those who have a DB
pension. There have been 9 waves of ELSA data collection -- the first was in
2001 and collections happened every two years thereafter.

Importantly, ELSA also includes information on pension size calculated by the
Institute for Fiscal Studies (IFS). This is only available up to and including
wave 5 (which was collected in 2010 and 2011) at which point I use a real return
of 3\% to predict forward the pension wealth variable until retirement. ELSA
also includes a measure of all non-housing financial wealth which I use in some
specifications because of these issues with pension wealth data. Due to slight
differences in the ELSA questions between years, I use `Harmonized ELSA'
\footnote{This analysis uses data or information from the Harmonized ELSA
    dataset and Codebook, Version G.2 as of July 2021 developed by the Gateway to
    Global Aging Data. The development of the Harmonized ELSA was funded by the
    National Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more
    information, please refer to https://g2aging.org/.} which ensures that variables
are comparable across waves. Since this only includes a subset of the questions
in ELSA, I supplement it with variables taken directly from the data such as
questions that deal with life expectancy.

@@

I also use `life tables' from the UK's ONS. These provide me with the risk of
death for each age group. Since the tables are produced until age 120, I adjust
them to make death certain at age 110 as is common in the literature
\cite{odea_sturrock_rest_2023}. I transform these so that I have risk of death
conditional on being a given age since this is what I use in the life cycle
simulations. I also use these objective probabilities to calculate annuity
prices for individuals.

To illustrate the effect that the reform had on sales of annuities in the UK I
obtained product data from the Financial Conduct Authority. These track the
sales of different financial products overtime including data on annuities.

Life expectancy impacts the decision to annuitise for two reasons. Firstly, it
directly impacts the price that an annuity will cost for individuals. Older
individuals and those with pre-existing health conditions generally can buy an
annuity that provides a greater income stream than individuals who are younger.
Secondly, private information about life expectancy impacts the perceived value
of an annuity. If an individual expects to outlive the general population, an
annuity, at a given price, will appear a much better deal to them. Likewise, an
individual's life expectancy will change their decision on how to consume and
save through retirement.

To calculate subjective life expectancies from ELSA data, I follow
\cite{odea_sturrock_rest_2023}. Individuals were asked “What are the chances
that you will live to be age X or more?” where X changed depending on the age of
the interviewee. If individuals were under 65 then X was 75, if individuals were
66 and older they were asked the age that was 11 to 15 years older than them and
a multiple of 5. From wave 3 respondents were also asked “What are the chances
that you will live to be age 85 or more?” if they were under 70. As most recent
retirees are under 70 we therefore have two data points. I first drop from the
data individuals who think it is more likely that they reach a higher age than a
younger age since this shows a misunderstanding of the question. I then add as a
third data point their objective chance of reaching 110 according to the ONS
life tables. I fit these three points to a Weibull distribution, which is
commonly used by demographers to estimate how populations will age, using
non-linear least squares. Then I create subjective survival tables using
parameters from the Weibull distribution.

@@

Successive governments in the UK have attempted to reform both the public and
private pension system. Prior to 1987, participation in private schemes was
limited to employees of firms that had offered them, and there were few
alternatives to the public state pension or DB scheme that a public sector
employer would offer. From 1989, individuals in the UK were allowed to open tax
advantaged self-invested personal pensions alongside any pension their employer
offered. The 2004 Finance Act rationalised taxation rates on DB and DC pensions.
The rules for DC pension pots were such that individuals had to buy an annuity
after optionally taking a maximum of 25\% of the pot as a tax free lump sum
withdrawal. DC pension pots were accessible from age 55 and most required that
they be accessed before the age of 75.

In the June 2010 budget, the coalition government made the first of two key
reforms it would make to the pension system between 2010 and 2015. The 2010
policy reform created a minimum income requirement above which individuals would
not need to annuitise more \cite{finance_act_hmt_2011}, this meant high income
retirees would not need to annuitise their wealth. The income requirement was
set at £20,000. This meant, for example, that an individual who was a member of
a DB scheme paying them £20,000 a year would not need to annuitise their DC
pension pot. However, the relatively high minimum income requirement meant that
most individuals still had to purchase an annuity.

In spring 2014, George Osbourne announced the so called `pensions freedom act'
scrapping the minimum income requirement and eliminating the compulsory
annuities market \cite{pen_freedoms_hmt_2014}. One government minister
infamously brought the message home by saying pension pots `can be used to buy
Lamborghinis' if retirees wanted to \cite{guardian_lambos}. The government also
announced plans to make switches between DB and DC pension schemes possible
although this was only realised in the next parliament -- future research could
explore the impact of this second reform on the consumption and saving decisions
of retirees.

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011. Figure \ref{fig:annovertime} shows purchases of annuities overtime and
demonstrates the sharp decrease in purchases that happened from 2014 to 2015.
There was an increase in the number of pensions that were being accessed using
an income drawdown product, but these only partly account for the drop in
annuitisation as is shown by the graph.
\begin{figure}[h]
    \caption{New pension pots accessed }
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
\end{figure}

\begin{figure}[h]
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_pot_sizes.pdf}
    \label{fig:ann2122}
    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution methods to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most defined contribution pensions were accessed through
annuities.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if it was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK.
However, given that this change was implemented two years before the pension
freedoms act it will not impact my results.

\footnotetext{Partial UFPLS (uncrystallised funds pension lump sum) is similar
    to Partial drawdown but with a different tax schedule. With partial drawdown
    you take the whole 25\% tax free amount at once whereas with UFPLS you only
    claim the tax relief on 25\% of the amount you are taking out. This option
    is better if you may want to buy another retirement product in the future
    such as an annuity. }

@@

Table \ref{tab:sum_stats}\footnote{There are hyperlinks on the table number so,
    if your pdf viewer supports this, you can click to go to the table or
    figure} shows various summary statistics for each variable of interest from
ELSA and the ONS.

As required, retirement year is between 2011 and 2013 for the control group and
2015 and 2017 for the treatment group. However, as discussed above their ELSA
interview date is after or the same as the retirement year because we are
tracking consumption in retirement.

@@

The treatment group has higher financial wealth with a median of £66,000 at the
time of interview as opposed to £55,000 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
much more money in them. Both groups are equally likely to have a DB pension
with roughly 44\% of individuals across both samples having a DB pension. In
general, DB pensions are more prevalent than DC pensions in the data, this
reflects the same trend that is observed in the general UK population. Home
ownership is roughly equal across groups although the second group have slightly
higher housing wealth.

@@

\textbf{Maybe add in prob of survival to age 85 or 90 or something}
\textbf{Also add in an indicator for living alone or living with someone else}

Overall, the groups appear to have slightly different wealth profiles with the
treatment group being richer and slightly more likely to have DC pension wealth.
On key demographic characteristics, however, the groups are similar. The
difference in retirement age between the two groups can be mostly explained by
the difference in expected retirement age meaning that individuals have not en
masse decided to delay retirement and avoid annuitisation. Moreover, for the
regression discontinuity models I only use individuals who have a DC pension so
it does not matter that there are more individuals with DC pensions in the
second group.
\begin{landscape}
    \linespread{1.25}
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

As a further test for balance I regress demographic and financial
characteristics of individuals on the treatment dummy. We can then see if the
treatment groups differ on key characteristics such as financial wealth or
retirement age. For a regression discontinuity to be valid we need the treatment
and control groups to be similar along all other characteristics apart from the
treatment variable.

@@

Where $Y_{i}$ are the different demographic and financial characteristics.

\linespread{1.25}
\input{tables/cov_balance.tex}
\linespread{1.5}

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0 which shows that individuals are not
delaying retirement in order to be part of the treatment group although
retirement age is higher post reform. DC pension value is higher, as is
financial wealth but these are both imprecisely estimated. Because of the
differences in these variables I add them to the regression model, following

One threat to validity is manipulation of the running variable, retirement year.
This is probably quite a threat in this setting as individuals could delay
retirement and/or the purchase of an annuity until after the reform. As observed
above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it does not
appear like individuals are delaying retirement. However, it is possible that
individuals do not buy an annuity at the time of retirement since the law prior
to 2014 only required that they access the pot by age 75. They may decide to
keep their DC pension untouched and live off other income and assets before
accessing it later. Since we do not have data on the exact purchase date of
annuities we cannot track whether annuity purchases were delayed because of
this. However, there is relatively stable annuity demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so they would not have to buy an annuity at all, we would observe declines in
purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a big spike in annuity purchases prior to
2014.

@@

In this section, I outline the key empirical models I run with both the simulated
consumption data from the lifecycle models and with the real data from ELSA. I
then see which lifecycle model better fits the consumption response that
occurred as a result of the pension reform.

I use a fuzzy regression discontinuity design for which I compare the
consumption of individuals who retired after the policy reform to that of people
who retired just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing the
change as a ``bombshell" \cite{money_marketing_announcement}.

>>>>

\section{Introduction}
Classical economic theory suggests that retirement annuities, which provide a
guaranteed income stream until the end of life, should be highly valued by
individuals as a way to insure against late death \citep{yaari_65}. However, in
developed countries, rates of annuitisation are far below the levels that theory
predicts -- this phenomenon has been termed the ``annuity puzzle". The literature
has suggested several reasons for this, but there is no consensus about which
mechanism is dominant. I exploit early retirees' consumption responses to a
reform to annuity policy in the UK to add to this literature. [\textbf{Add
            something once I have a conclusion}].

In the UK, adults have three methods of funding retirement: the state pension;
defined benefit (DB) employee pension schemes, which provide an income in
retirement based on years of service and a function of wages; and defined
contribution (DC) pensions, under which individuals save and invest in a
tax-advantaged account that is usually supplemented with employer contributions.
Under the 2010-2015 coalition government in the UK, the law regarding the use of
private DC pensions at retirement changed. Individuals were no longer forced to
annuitise their pension pots and could access them in a variety of ways, such as
a lump sum withdrawal or income drawdown, involving a steady withdrawal of
assets from the pension pot -- common advice is to take 4\% a year -- whilst the
rest remains invested. Subsequently, the number of annuities sold in the UK
dropped precipitously.

In this paper, I first use the policy reform as a discontinuity to measure the
impact of forced annuitisation on the consumption levels of individuals in the
first few years of retirement. Given that the reform was implemented suddenly
and without advanced notice before the Spring 2014 budget, I claim that
individuals who retired in 2015, 2016 and 2017 are otherwise similar to those
who retired in 2011, 2012 and 2013 except for the fact that the later retirees
do not need to annuitise their DC pension pots. Therefore, I can use a
regression discontinuity to identify the effect that forced annuitisation has on
the consumption of retirees.

I then test two competing hypotheses for the annuity puzzle: bequests and
pessimistic life expectancy. The annuity puzzle was first identified by
\cite{yaari_65} and is a classed as such since standard economic theory cannot
explain the lack of annuitisation that is observed in the real world. Two
possible explanations are bequests and life expectancy: individuals want to
leave an estate for their heirs to inherit and they cannot do this if they
annuitise their wealth; or individuals may believe that they will not live as
long as annuity providers think they will, and therefore annuities that are on
the market do not appear to be good value. Depending on the reason for the lack
of annuitisation in the UK, the consumption response of retirees to the pension
reform will differ. If individuals do not annuitise because of pessimistic life
expectancy, their consumption should increase after the reform. If, on the other
hand, individuals do not annuitise because of a bequest motive, consumption
should not change as much as a result of the reform.

I will solve lifecycle models for both of these cases and simulate consumption
decisions with and without forced annuitisation. I will then use a variety of
empirical models to measure the consumption change in early retirement that
resulted from the policy reform. The sign and magnitude of this change will be
indicative of the mechanism causing the annuitisation problem.

The importance of retirement policy to individuals in the UK is growing. The
number of individuals of pensionable age is expected to increase from 11.9
million in 2020 to 15.2 million in 2045 according to the latest projections from
the Office for National Statistics (ONS), and for every 1,000 people of working
age there will be 341 of pensionable age in 2045 compared to 280 in 2020
\citep{ons_population_predictions_2020}. The increase in absolute and relative
numbers of elderly people makes retirement policy more consequential. Moreover,
private DC pensions are becoming increasingly common and are predicted to grow
as current cohorts age \citep{cribb_karjalainen_ifs_2023}. Therefore, policies
regarding how private pensions can be accessed will have a larger impact on
overall welfare for retirees.

Moreover, understanding how retirees spend their money over retirement is an
important policy question of its own. If retirees spend too much in early
retirement, the state may need to provide for them towards the end of their
lives. This has implications for government budgets, especially since population
aging means more individuals will require expensive end of life care. On the
other hand, if retirees over-save and do not spend, the economy may be
dynamically inefficient -- when the capital stock is too high -- because
individuals save too much for retirement and for bequests. In this case
consumption per capita could be increased if savings were decreased.

\subsection{Literature review}
My paper draws on three main strands of literature: that of the annuity puzzle,
exploring reasons why retirees choose not to annuitise; the retirement
saving puzzle, which seeks to explain why retirees drawdown assets slowly; and
lifecycle models, whereby the above problems are explained using a tractable
model of human decision making.

\cite{yaari_65} showed that under standard assumptions with risk averse agents
we would expect individuals to annuitise all of their wealth at retirement to
insure against the risk of long life. And more recently
\cite{davidoff_brown_diamond_aer_2005} show that complete annuitisation is
optimal under a broader range of assumptions. Since then there has been much
discussion of possible reasons for why people do not annuitise.
\cite{finkelstein_porteba_2002, finkelstein_porteba_2004} find evidence of
adverse selection into the UK annuity market and \cite{mitchell_et_al_1999} find
evidence of the same in the US. This means longer lived individuals select into
buying annuities which drives up annuity prices because providers need to pay
individuals for longer than they originally thought they would need to. This
makes the `money's worth' of an annuity lower for the general population as
opposed to the population of annuitants, putting off individuals from buying an
annuity. However, \cite{mitchell_et_al_1999} also find that theory would still
predict annuitisation because the money's worth of annuities is not that far
from `actuarially fair' (this is the where the present discounted value of annuities
equals the price that an individual pays for it).

\cite{friedman_warshawsky_qje_1990} show that annuitisation decisions can be
fully explained by a mixture of bequest motives and `actuarially unfair'
annuities. They solve an augmented life-cycle model and calculate annuity demand
for a range of differences between rate of return assets would get in the stock
market and the rate of return annuities give an individual. For plausible values
they find that individuals would optimally not annuitise much wealth. Similarly
to \cite{finkelstein_porteba_2004}, \cite{friedman_warshawsky_chicago_1988} show
that there is a significant difference between the life expectancy of annuitants
and the general population in the American annuity market, which therefore
implies a degree of adverse selection. But they find this cannot fully explain
the annuitisation puzzle and only when bequest motives are added to the model
can annuitisation rates be rationalized.

More recently, \cite{lockwood_red_2012} builds on this and shows that a
realistic bequest motive in lifecycle simulations achieves realistic
annuitisation rates. He solves a simple lifecycle model with bequest motives
taken from several recent papers in the retirement savings literature. The
bequest motives he picks matches other important aspects of empirical
distribution of bequests such as how much individuals give and the financial
characteristics of these individuals. Also, \cite{odea_sturrock_rest_2023} find
that subjective life expectancies can explain a large part of the annuitisation
puzzle especially if agents are less risk averse.

Apart from bequests and annuity prices, a range of other reasons for the annuity
puzzle have been explored. Pre-cautionary savings for unexpected end-of-life
medical expenses might cause individuals to prefer holding assets rather than a
an annuity. Moreover, \cite{vidalmelia_lejarragagarcia_munich_2004} suggests
that couples might share the risk of living for a long-time between them and
therefore couples may be less likely to annuitise their wealth.

Given that uncertain end-of-life medical expenses are not such an issue in the
UK as opposed to the US, and that the focus of recent literature has been on the
bequest model and uncertain life expectancy, I choose also to focus on bequest
and objective life expectancy in the UK context.

@@

The main dataset I use is the English Longitudinal Study of Ageing (ELSA)
\citep{main_elsa_citation}. ELSA picks individuals over the age of 50 to survey
every two years until death. If individuals leave the survey, ELSA replaces them
so that it is representative of the over 50 population in the UK. Individuals
are asked a range of questions relating to their income and wealth as well as
expectations about the future. One benefit of using ELSA is that it includes
data on pension types for individuals who are working. Therefore, I can
differentiate between individuals who have a DC pension and those who have a DB
pension. There have been 9 waves of ELSA data collection -- the first was in
2002/03, and collections happened every two years thereafter.

Importantly, ELSA also includes information on pension size calculated by the
Institute for Fiscal Studies (IFS). This is only available up to and including
wave 5 (which was collected in 2010 and 2011), after which point I use a real return
of 3\% to predict forward the pension wealth variable until retirement. ELSA
also includes a measure of all non-housing financial wealth which I use in some
specifications because of these issues with pension wealth data. Due to slight
differences in the ELSA questions between years, I use `Harmonized ELSA'
\footnote{This analysis uses data or information from the Harmonized ELSA
    dataset and Codebook, Version G.2 as of July 2021 developed by the Gateway to
    Global Aging Data. The development of the Harmonized ELSA was funded by the
    National Institute on Aging (R01 AG030153, RC2 AG036619, R03 AG043052). For more
    information, please refer to https://g2aging.org/.} which ensures that variables
are comparable across waves. Since this only includes a subset of the questions
in ELSA, I supplement it with variables taken directly from the data such as
questions that deal with life expectancy.

@@

I also use `life tables' from the UK's ONS from 2014. These provide me with the
risk of death for each age group. The tables are produced until age 120 and I
adjust them to make death certain at age 110 as is common in the literature
\citep{odea_sturrock_rest_2023}. I transform these so that I have risk of death
conditional on being a given age since this is what I use in the life cycle
simulations. I also use these objective probabilities to calculate annuity
prices for individuals.

To illustrate the effect that the reform had on sales of annuities in the UK, I
obtained product data from the Financial Conduct Authority (FCA). These track the
sales of different financial products over time including data on annuities.

Life expectancy impacts the decision to annuitise for two reasons. Firstly, it
directly impacts the price that an annuity will cost for individuals. Older
individuals and those with pre-existing health conditions generally can buy an
annuity that provides a greater income stream than individuals who are younger.
Secondly, private information about life expectancy impacts the perceived value
of an annuity. If an individual expects to outlive the general population, an
annuity, at a given price, will appear a much better deal to them. Likewise, an
individual's life expectancy will affect their decisions over how to consume and
save through retirement.

To calculate subjective life expectancies from ELSA data, I follow
\cite{odea_sturrock_rest_2023}. Individuals were asked “What are the chances
that you will live to be age X or more?” where X changed depending on the age of
the interviewee. If individuals were under 65 then X was 75; if individuals were
66 and older they were asked the age that was 11 to 15 years older than them and
a multiple of 5. From wave 3, respondents were also asked “What are the chances
that you will live to be age 85 or more?” if they were under 70. As most recent
retirees are under 70 we therefore have two data points. I first drop from the
data individuals who think it is more likely that they reach a higher age than a
younger age since this shows a misunderstanding of the question. I then add as a
third data point their objective chance of reaching 110 according to the ONS
life tables. I fit these three points to a Weibull distribution, which is
commonly used by demographers to estimate how populations will age, using
non-linear least squares. Then I create subjective survival tables using
parameters from the Weibull distribution.

@@

Successive governments in the UK have attempted to reform both the public and
private pension system. Prior to 1987, participation in private schemes was
limited to employees of firms that had offered them, and there were few
alternatives to the public state pension or DB scheme that a public sector
employer would offer. From 1989, individuals in the UK were allowed to open tax
advantaged self-invested personal pensions alongside any pension their employer
offered. The 2004 Finance Act rationalised taxation rates on DB and DC pensions.
The rules for DC pension pots were such that an individual had to buy an annuity
after optionally taking a maximum of 25\% of the pot as a tax-free lump sum
withdrawal. DC pension pots were accessible from age 55, and most required that
they be accessed before the age of 75.

In the June 2010 budget, the coalition government announced the first of two key
reforms it would make to the pension system between 2010 and 2015. The 2010
policy reform created a minimum income requirement above which individuals would
not need to annuitise \citep{finance_act_hmt_2011}, meaning high income
retirees would not need to annuitise their wealth. The income requirement was
set at £20,000. This meant, for example, that an individual who was a member of
a DB scheme paying them £20,000 a year would not need to annuitise their DC
pension pot. However, the relatively high minimum income requirement meant that
most individuals still had to purchase an annuity.

In spring 2014, Osbourne, the chancellor at the time, announced the so called
`pensions freedom act', scrapping the minimum income requirement and eliminating
the compulsory annuities market \citep{pen_freedoms_hmt_2014}. One government
minister infamously brought the message home by saying pension pots `can be used
to buy Lamborghinis' if retirees wanted to \citep{guardian_lambos}. The
government also announced plans to make switches between DB and DC pension
schemes possible although this was only realised in the next parliament.

\textbf{For conclusion: future research could explore the impact of this second reform on the
    consumption and saving decisions of retirees.}

The impact of the reforms on annuity demand has been documented by
\cite{cannon_et_al_nier_2016}. Using data from the Association of British
Insurers, they show that annuity demand dropped by 75\% from its maximum in
2011 to 2016. Figure \ref{fig:annovertime} shows
purchases of annuities over time and demonstrates the sharp decrease in purchases
that ocurred from 2014 to 2015. There was an increase in the number of pensions
that were being accessed using an income drawdown product, but these only partly
account for the drop in annuitisation as is shown by the graph.
\begin{figure}[h]

    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_overtime.pdf}
    \label{fig:annovertime}
    \caption{New pension pots accessed}
\end{figure}

\begin{figure}[h]

    \centering
    \includegraphics[width=0.9\columnwidth]{figures/annuity_pot_sizes.pdf}
    \caption[Caption for LOF]{How pension pots are accessed\protect\footnotemark}
    \label{fig:ann2122}

    \
\end{figure}
Figure \ref{fig:ann2122} shows the distribution of methods used to access pension pots
at retirement in 2021-22. In 2021, 196,736 pots were fully withdrawn at
retirement, accounting for over 50\% of pots. Prior to the policy reform this
was not the case -- most DC pensions were accessed through
annuities.

Also of specific interest to the annuity market was the European Union's `Equal
Treatment in Goods and Services Directive' of 2004. This prohibited
discrimination in the provision and cost of goods and services based on sex. Up
until 2011, there was a clause that stated insurers were allowed to charge
different premiums if this was based on evidence that sex is correlated with
different amounts of risk. However, in March 2011, the European Court of Justice
ruled that insurers were not allowed to charge different amounts and gave them
until December 2012 to implement the ruling. This change meant that annuity
products could no longer be priced differently for men and women in the UK.
However, given that this change was implemented two years before the pension
freedoms act it will not impact my results.

\footnotetext{Partial UFPLS (uncrystallised funds pension lump sum) is similar
    to Partial drawdown but with a different tax schedule. With partial drawdown
    you take the whole 25\% tax-free amount at once whereas with UFPLS you only
    claim the tax relief on 25\% of the amount you are taking out. This option
    is better if you may want to buy another retirement product in the future
    such as an annuity. }

@@

Table \ref{tab:sum_stats}\footnote{There are hyperlinks on the table number so,
    if your pdf viewer supports this, you can click to go to the table or
    figure.} shows various summary statistics for each variable of interest from
ELSA and the ONS.

As required, retirement year is between 2011 and 2013 for the control group and
2015 and 2017 for the treatment group. However, as discussed above their ELSA
interview date are after or the same as the retirement year because we are
tracking consumption in retirement.

@@

The treatment group has higher financial wealth with a median of £67,000 at the
time of interview as opposed to £55,000 in the control group. Likewise, the
second group are more likely to have held a DC pension at some point and have
more money in them. Both groups are equally likely to have a DB pension with
roughly 45\% of individuals across both samples having a DB pension. In general,
DB pensions are more prevalent than DC pensions in the data -- this reflects the
trend that is observed in the general UK population. Home ownership is roughly
equal across groups although the second group has slightly higher housing
wealth.

@@

Overall, the groups appear to have slightly different wealth profiles with the
treatment group being richer and slightly more likely to have DC pension wealth.
On key demographic characteristics, however, the groups are similar. The
difference in retirement age between the two groups can be mostly explained by
the difference in expected retirement age meaning that individuals have not en
masse decided to delay retirement and avoid annuitisation. Moreover, for the
regression discontinuity models I only use individuals who have a DC pension, so
it does not matter that there are more individuals with DC pensions in the
second group.

\textbf{conclusion: chat about couples, chat about better data on at what
    stage of retirement individuals buy annuities}

\begin{landscape}
    \linespread{1.25}
    \input{tables/r_summary_stats.tex}
    \normalsize
\end{landscape}

As a further test for balance I regress demographic and financial
characteristics of individuals on the treatment dummy. We can then see if the
treatment and control groups differ on key characteristics such as financial
wealth or retirement age. For a regression discontinuity to be valid we need the
groups to be similar along all other characteristics apart from the treatment
variable.

@@

Where $Y_{i}$ are the different demographic and financial characteristics for
only those individuals with a DC pension.

\linespread{1}
\input{tables/cov_balance.tex}
\linespread{1.5}

Table \ref{tab:cov_balance} shows these results. As expected, retirement year
and interview year are greater for the treated group. The difference between
expected and real retirement age is 0 which shows that individuals are not
delaying retirement, beyond their original expectation,  in order to be part of
the treatment group although retirement age is higher post reform. DC pension
value is higher, as is financial wealth but these are both imprecisely
estimated. Because of the differences in these variables I add them to the
regression model.

One threat to validity is manipulation of the running variable, retirement year.
This is probably quite a threat in this setting as individuals could delay
retirement and/or the purchase of an annuity until after the reform. As observed
above in Tables \ref{tab:sum_stats} and \ref{tab:cov_balance}, it does not
appear as though individuals are delaying retirement. However, it is possible
that individuals do not buy an annuity at the time of retirement since the law
prior to 2014 only required that they access the pot by age 75. They may decide
to keep their DC pension untouched and live off other income and assets before
accessing it later. Since we do not have data on the exact purchase date of
annuities, we cannot track whether annuity purchases were delayed because of
this. However, there is relatively stable annuity demand before the policy
reform. If individuals delayed annuity purchases in expectation of the reform,
so that they would not have to buy an annuity at all, we would observe declines
in purchases before the reform was announced. Likewise, if individuals bought an
annuity early in expectation that the end of the compulsory market would cause
prices to go up, we would observe a spike in annuity purchases prior to 2014.

@@

In this section, I outline the key empirical models I run with first the real
data from ELSA, and then the simulated data from the lifecycle models. I then
see which lifecycle model better fits the consumption response that occurred as
a result of the pension reform.

I use a fuzzy, since some individuals will probably still annuitise a little,
regression discontinuity design for which I compare the consumption of
individuals who retired after the policy reform to that of people who retired
just before the reform. The key assumption implicit in regression
discontinuities is that nothing else changes at the time of the jump apart from
the policy of interest, and that the policy occurs without individuals
predicting it and altering their behaviour. As I have argued above, the
demographic information in the data suggests that individuals did not delay
retirement and that there was no delay in annuitisation as shown by the quick
and sudden decline in annuity purchases. Moreover, anecdotal evidence from media
and business sources at the time of the announcement show that there was
surprise the government put forward the reform, with Money Marketing describing
the change as a ``bombshell" \citep{money_marketing_announcement}.
